---
title: 'Reproducible Research: Peer Assessment 1'
author: "Carlos Guadián"
date: "12/2/2021"
output: 
  html_document:
    keep_md: yes
    toc: yes

---

```{r setup}
knitr::opts_chunk$set(echo = TRUE)
```

## Load & process data
First load data into `activity` variable
```{r read data}
activity <- read.csv("~/RepData_PeerAssessment1/activity.csv")
```

Check for correct class of variables
```{r check data again}
str(activity)
```

`date` variable has class character, and it has to be "Date".
```{r convert date}
activity$date <- as.Date(activity$date)
```

Check again
```{r check data}
str(activity)
```

## What is mean total number of steps taken per day?
**Calculate the total number of steps taken per day**

To calculate the total number of steps taken per day first I have to group data by day. To do this we need the `tidyverse` library.
```{r group data}
library(tidyverse)
steps_by_day <- activity %>%
  group_by(date) %>%
  summarize(total_steps = sum(steps))
head(steps_by_day)
```

**Make a histogram of the total number of steps taken each day**

Making histogram with base system
```{r histogram,fig.height=5,fig.cap="Histogram of total steps by day"}
with(steps_by_day, {
  hist(total_steps, col = "green", main = "Histogram total steps by day", xlab = "Total steps")
})
```

**Calculate and report the mean and median of the total number of steps taken per day**
Using two ways to get mean and median. With `summary` and `mean()` & `median()`
```{r mean median}
summary(steps_by_day$total_steps)
mean(steps_by_day$total_steps, na.rm = TRUE)
median(steps_by_day$total_steps, na.rm = TRUE)
```

## What is the average daily activity pattern?
**Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)**

Group by interval to get average number of steps by interval
```{r group interval, echo= TRUE}
steps_by_interval <- activity %>%
  group_by(interval) %>%
  summarize(average_steps = mean(steps, na.rm = TRUE))
head(steps_by_interval)
```

Plotting time serie
```{r time serie}
with(steps_by_interval, {
plot(interval, average_steps, type = "l",
     main = "Average number of steps by 5 min. interval", xlab = "5 minutes Interval", ylab = "Average Steps", col = "blue", lwd = 1, frame.plot=FALSE, xaxt="n")
  axis(1, at = seq(00, 2355, by = 50), las=2)
  })
```

**Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?**
Getting the interval filtering with `which.max()`
```{r interval max}
max_steps_interval <- steps_by_interval[which.max(steps_by_interval$average_steps),]
max_steps_interval
```
Adding vertical line to check if interval 835 has the maximum steps in graph
```{r time serie max,fig.cap="Average number of steps by 5 min. interval"}
with(steps_by_interval, {
plot(interval, average_steps, type = "l",
     main = "Average number of steps by 5 min. interval", xlab = "5 minutes Interval", ylab = "Average Steps", col = "blue", lwd = 1, frame.plot=FALSE, xaxt="n")
  axis(1, at = seq(00, 2355, by = 50), las=2)
  abline(v=835, col = "red", lwd = 1)
  })
```

## Imputing missing values
Note that there are a number of days/intervals where there are missing values (**coded as NA**). The presence of missing days may introduce bias into some calculations or summaries of the data.

**Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)**
The only variable in `activity` dataset with NAs is `steps`, we can check this with `summary`, and then we can confirm with `sum(is.na())`
```{r missing values}
summary(activity)
sum(is.na(activity$steps))
```

**Create a new dataset that is equal to the original dataset but with the missing data filled in.**

**Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.**

To replace NA we get the mean of each interval and then substitute the NA for this new data.
```{r replace NA}
library(dplyr)
# Create new dataset to fill NA values
intervals <- activity
# Groupping and replacing
intervals <- intervals %>%
  left_join(intervals %>%
              group_by(interval) %>%
              summarise(mean_steps = mean(steps, na.rm = TRUE))) %>%
  mutate(steps = ifelse(is.na(steps), mean_steps, steps)) %>%
  select(-mean_steps)
```

Check if the replace are OK
```{r check NA}
# First if there are NA values
summary(intervals)
sum(is.na(intervals$steps))
```

**Make a histogram of the total number of steps taken each day**

To calculate the total number of steps taken per day first I have to group data by day. To do this we need the `tidyverse` library.
```{r histogram steps no NA}
steps_by_day_clean <- intervals %>%
  group_by(date) %>%
  summarize(total_steps = sum(steps))
head(steps_by_day_clean)
```

Making histogram
```{r histogram no NA}
with(steps_by_day_clean, {
  hist(total_steps, col = "green", main = "Histogram total steps by day without NA", xlab = "Total steps")
})
```

**Calculate and report the mean and median total number of steps taken per day**
```{r mean median no NA}
summary(steps_by_day_clean$total_steps)
mean(steps_by_day_clean$total_steps, na.rm = TRUE)
median(steps_by_day_clean$total_steps, na.rm = TRUE)
```

**Do these values differ from the estimates from the first part of the assignment?**

Comparing two values
```{r compare}
#comparing mean
identical(mean(steps_by_day$total_steps, na.rm = TRUE), mean(steps_by_day_clean$total_steps, na.rm = TRUE))
# comparing median
identical(median(steps_by_day$total_steps, na.rm = TRUE),median(steps_by_day_clean$total_steps, na.rm = TRUE))
```

**What is the impact of imputing missing data on the estimates of the total daily number of steps?**

```{r summarizing}
# Summarizing
# With NA
summary(steps_by_day)
#Without NA
summary(steps_by_day_clean)
```


## Are there differences in activity patterns between weekdays and weekends?

For this part the `weekdays()` function may be of some help here. Use the dataset with the filled-in missing values for this part.

Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.
```{r factoring}
# In order to get correct running if you don't have english locale, first set up
Sys.setlocale(locale="en_US.UTF-8")
#create a vector of weekdays
weekdays1 <- c('Monday', 'Tuesday', 'Wednesday', 'Thursday', 'Friday')
#Use `%in%` and `weekdays` to create a logical vector
#convert to `factor` and specify the `levels/labels`
intervals$wDay <- factor((weekdays(intervals$date) %in% weekdays1), 
         levels=c(FALSE, TRUE), labels=c('weekend', 'weekday'))
table(intervals$wDay)
```

**Make a panel plot containing a time series plot type = "l" of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).**

Grouping by interval and wDay
```{r grouping wDay, message=FALSE, warning=FALSE}
weekday_vs_weekend <- intervals %>%
  group_by(interval, wDay) %>%
  summarise(steps_mean = mean(steps))
weekday_vs_weekend
```

Plotting panel to compare weedays with weekend
```{r}
library(lattice)
table(weekday_vs_weekend$wDay)
with(weekday_vs_weekend, {
  xyplot(steps_mean ~ interval | wDay, type = "l", layout = c(1,2), main = "Comparing steps between weedays and weekends")
})
```
```That's all```